Modelling last glacial cycle ice dynamics in the Alps

### 2017-11-20 articles/cycle/alps-cycle.tex
Add reference to Haeberli and Schluchter (1987).

    The spatial extent and thickness of Alpine glaciers during the LGM, an
    integrated footprint of thousands of years of climate history and glacier
    dynamics, have been reconstructed from moraines and trimlines across the
    mountain range \citep[e.g.,][]{Bini.etal.2009, Coutterand.2010,
    Husen.2011}. The assumption that trimlines, the upper limit of glacial
    erosion, represent the maximum ice surface elevation, has
    been repeatedly invalidated in other glaciated regions of the globe
    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012}. Yet no
    evidence for cold-based glaciation has been reported in the Alps to date.
    Alpine ice flow
    patterns were primarily controlled by subglacial topography, but there is
    evidence for flow accross major mountain passes
    \citep[e.g.,][]{Coutterand.2010, Kelly.etal.2004, Husen.2011}, and
    self-sustained ice domes \citep{Bini.etal.2009}. Finally, the timing of the
    LGM in the Alps \citep{Ivy-Ochs.etal.2008, Monegato.etal.2017} is in good
    agreement with the maximum expansion of continental ice sheets recorded by
    the Marine Oxygen Isotope Stage (MIS) 2
    \citep[29--14\,ka;][]{Lisiecki.Raymo.2005}. Regional variation between
    different piedmont
    lobes exists \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear whether it
    relate to climate or glacier dynamics \citep{Monegato.etal.2017} or
    uncertainties in the dating methods.

@@

    In the upper Rhone valley, the geographical location of observed trimlines
    show a remarkable agreement with the areal extent of regions that are
    modelled to have experienced warm-based ice for less than 1\,ka (7\,\%) of the
    MIS~2 (Fig.~\ref{fig:trimlines}c). Although this result calls for more
    detailed comparisons across the
    entire Alpine range and specific sensitivity studies to relevant basal
    sliding and ice rheological parameters, and to the uncertain subglacial
    topography the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable. In this context, our results challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer of cold ice in the Alps.

>>>>

    The spatial extent and thickness of Alpine glaciers during the LGM, an
    integrated footprint of thousands of years of climate history and glacier
    dynamics, have been reconstructed from moraines and trimlines across the
    mountain range \citep[e.g.,][]{Bini.etal.2009, Coutterand.2010,
    Husen.2011}. The assumption that trimlines, the upper limit of glacial
    erosion, represent the maximum ice surface elevation, has
    been repeatedly invalidated in other glaciated regions of the globe
    \citep[e.g.,][]{Kleman.1994, Kleman.etal.2010, Fabel.etal.2012}. Although
    no evidence for cold-based glaciation has been reported in the Alps to
    date, it is supported by numerical modelling
    \citep[Fig.~3]{Haeberli.Schluchter.1987}. Alpine ice flow
    patterns were primarily controlled by subglacial topography, but there is
    evidence for flow accross major mountain passes
    \citep[e.g.,][]{Coutterand.2010, Kelly.etal.2004, Husen.2011}, and
    self-sustained ice domes \citep{Bini.etal.2009}. Finally, the timing of the
    LGM in the Alps \citep{Ivy-Ochs.etal.2008, Monegato.etal.2017} is in good
    agreement with the maximum expansion of continental ice sheets recorded by
    the Marine Oxygen Isotope Stage (MIS) 2
    \citep[29--14\,ka;][]{Lisiecki.Raymo.2005}. Regional variation between
    different piedmont
    lobes exists \citep[Fig.~5]{Wirsig.etal.2016}, but it is unclear whether it
    relate to climate or glacier dynamics \citep{Monegato.etal.2017} or
    uncertainties in the dating methods.

@@

    In the upper Rhone valley, the geographical location of observed trimlines
    show a remarkable agreement with the areal extent of regions that are
    modelled to have experienced warm-based ice for less than 1\,ka (7\,\%) of the
    MIS~2 (Fig.~\ref{fig:trimlines}c). Although this result calls for more
    detailed comparisons across the
    entire Alpine range and specific sensitivity studies to relevant basal
    sliding and ice rheological parameters, and to the uncertain subglacial
    topography the presence of an upper layer of
    cold ice during the LGM, already found at high altitude in the much warmer
    climate of today \citep[e.g.,][]{Suter.etal.2001, Bohleber.etal.2017}, is
    inevitable \citep[Fig.~3]{Haeberli.Schluchter.1987}.
    In this context, our results challenge the assumption that
    Alpine trimlines mark the maximum upper ice surface elevation of the LGM
    ice cover and call for a more accurate estimation of the thickness of the
    upper layer of cold ice in the Alps.
